Efficient pedigree recording for fast population genetics simulation

### 2018-01-15 forwards_paper.tex
minor smoothing

% I may need to cite my student's t-shirt for the following line:
%  "that's all very good in practice, but how does it hold up in theory?"
Figures \ref{fig:runtimes_selection} and \ref{fig:relative_speedup_selection} show that this method
can dramatically improve simulation performance in practice --
but, how does it perform in theory?
% The simulation results shown in Figures \ref{fig:runtimes_selection} and \ref{fig:relative_speedup_selection} show that
% pedigree tracking greatly speeds up forward-time simulations.  Further, the run times with pedigree tracking become
% near-linear in $\rho$ (Figure \ref{fig:runtimes_selection}).
Both computational time and storage space are depend mostly
on the number of \emph{mutations} and \emph{edges} in the tree sequence.
% Each edge represents a segment of genome
% that has been inherited across a number of generations equal to the length of the edge.
% Each of these generations provided
% an opportunity for mutation and for recombination
% (which changes the tree, thus usually introducing a new edge).
% For this reason, 
The key quantity to understand for this will be the total ``area'' of the tree sequence,
i.e., the product of segment length (left minus right coordinates)
and edge length (difference in birth times between parent and child),
summed across all edges.
This area is also equal to the sum of the total lengths of all marginal trees
(i.e., the trees describing inheritance at each position on the genome).
Therefore, the total area of a tree sequence is equal to the sequence length
multiplied by the mean area of a marginal tree.
Since we analyze tree sequences arising from a Wright--Fisher model,
statistical properties of a marginal tree are well-described by coalescent theory.
Similar arguments to those below go back at least to \citet{watterson1975number},
who also explicitly computed smaller order corrections relevant to whole-population genealogies
of the Wright--Fisher model.
For an introduction to coalescent theory, see \citet{wakeley2005coalescent}.

%% Memory reduction thanks to simplification
First: how much memory do simplified tree sequences require?
Consider a simulation of a Wright--Fisher population of $N$ haploid individuals
using Algorithm~\algref{W} for $T$ generations.
Since every chromosome inherits material from both parents,
without simplification this would produce tables of
$NT$ nodes and $2NT$ edges.
After simplification, we are left with the tree sequence describing the history
of only the current generation of $N$ individuals, back to either $T$ generations ago,
or their common ancestor, whichever comes first.
The tree sequence must store edges describing the leftmost marginal tree,
which requires at most $N-2$ edges.
Then, each time the marginal tree changes along the sequence,
four edges end and four new edges begin
(except for changes affecting the root, which require fewer; see \citet{kelleher2016efficient}).
Suppose that $T_x$ is the marginal tree describing how each of the present-day population
descends from their most recent common ancestor
at a particular genomic position $x$ on the chromosome,
and write $|T_x|$ for the total length of the tree,
i.e., the number of ancestors 
from which some but not all of the present-day population has inherited at position $x$
(including the population itself).
For the tree at nearby position $x + dx$ to be different,
there must have been a crossing-over between $x$ and $x + dx$
in one of the $|T_x|$ meioses that gave birth to those individuals.
(Recall that these ``individuals'' are haploid.)
Since we measure distance along the genome
so that length is equal to the expected number of crossing-overs per generation,
the expected distance until the next crossing-over is $1/|T_x|$.
If every crossing-over changed the marginal tree,
then this would imply, for consistency, 
that the expected total number of times that the marginal tree changes along the genome
is equal to the total area of the tree sequence (as defined above).
Not every such crossing-over changes the marginal tree, but most do,
so the total area of the tree sequence,
multiplied by four,
gives an upper bound on the expected number of edges in the tree sequence 
beyond those describing the leftmost tree.
Since we are considering a chromosome of length 1,
the expected total area is equal to the mean marginal tree length, as above.

If $T$ is large enough that all samples coalesce within the simulation with high probability
-- $T = 20N$, say --
then coalescent theory tells us that
the expected total length of the branches of a marginal tree back to the most recent common ancestor
is approximately $2N\log(N)$.
Therefore, the tree sequence describing the entire population
is expected to need no more than $2N + 8N\log(N)$ edges.
Not every new edge derives from a never-before-seen node,
but the number of nodes is at most equal to the number of edges plus the sample size.
Therefore, 
we would need $O(N^2)$ space to store the complete history of the simulation, 
but only $O(N \log N)$ % $O(N + N \log N)$
to store the history that is relevant to the extant population.

@@

%% Above was edges; same holds for memory usage savings due to mutations
What about mutations?
If we were to add mutations in forwards time
under the infinite-sites model with total mutation rate per generation $\mu$,
there would also be around $\mu N T$ mutations, and the same number of sites.
% simply because there was a total of $N T$ meioses.
Since mutations fall as recombinations do on the marginal trees,
the number of mutations is proportional to the total length of all the marginal trees,
and so by the same argument as for the number of edges,
adding them after simplification results in only around $\mu 2 N \log(N)$ mutations.
With $N=2 \times 10^4$ and $T=10N$,
this implies that adding neutral mutations afterwards reduces the number of mutations that must be generated
by a factor of 10,000.
This could result in substantial time savings,
even without considering the
computational burden of propagating mutations forwards across generations.

Since each mutation is stored only as a single row in the mutation table,
and at most one row in the site table, the space required for $M$ mutations is $O(M)$;
combined with the $O(N \log N)$ storage for edges and nodes of a simplified tree sequence above,
this implies that the full results of a simulation of $N$ individuals
having $M$ mutations can be stored in $O(N \log N + M)$ space.
In simulations of whole chromosomes, there are typically a small, bounded number of mutations per chromosome per generation,
so $M$ will be $O(N \log N)$ as well.

>>>>

% I may need to cite my student's t-shirt for the following line:
%  "that's all very good in practice, but how does it hold up in theory?"
Figures \ref{fig:runtimes_selection} and \ref{fig:relative_speedup_selection} show that this method
can dramatically improve simulation performance in practice --
but, how does it perform in theory?
% The simulation results shown in Figures \ref{fig:runtimes_selection} and \ref{fig:relative_speedup_selection} show that
% pedigree tracking greatly speeds up forward-time simulations.  Further, the run times with pedigree tracking become
% near-linear in $\rho$ (Figure \ref{fig:runtimes_selection}).
Both computational time and storage space are depend mostly
on the number of \emph{mutations} and \emph{edges} in the tree sequence.
% Each edge represents a segment of genome
% that has been inherited across a number of generations equal to the length of the edge.
% Each of these generations provided
% an opportunity for mutation and for recombination
% (which changes the tree, thus usually introducing a new edge).
% For this reason,
The key quantity to understand for this will be the total ``area'' of the tree sequence,
which is the sum of the lengths of all ancestral genomic segments that some, but not all,
of the present population has inherited.
It can be found by summing the product of segment length (left minus right coordinates)
and edge length (difference in birth times between parent and child),
across all edges.
This area is also equal to the sum of the total lengths of all marginal trees
(i.e., the trees describing inheritance at each position on the genome),
so can be computed as the sequence length multiplied by the mean marginal tree length.
Since we analyze tree sequences arising from a Wright--Fisher model,
statistical properties of a marginal tree are fairly well-described by coalescent theory.
Similar arguments to those below go back at least to \citet{watterson1975number},
who also explicitly computed smaller order corrections relevant to whole-population genealogies
of the Wright--Fisher model.
For an introduction to coalescent theory, see \citet{wakeley2005coalescent}.

%% Memory reduction thanks to simplification
First: how much memory do simplified tree sequences require?
Consider a simulation of a Wright--Fisher population of $N$ haploid individuals
using Algorithm~\algref{W} for $T$ generations.
Since every chromosome inherits material from both parents,
without simplification this would produce tables of
$NT$ nodes and $2NT$ edges.
After simplification, we are left with the tree sequence describing the history
of only the current generation of $N$ individuals, back to either $T$ generations ago,
or their common ancestor, whichever comes first.
The tree sequence must store edges describing the leftmost marginal tree,
which requires at most $N-2$ edges.
Then, each time the marginal tree changes along the sequence,
four edges end and four new edges begin
(except for changes affecting the root, which require fewer; see \citet{kelleher2016efficient}).
Suppose that $\mathcal{T}_x$ is the marginal tree at genomic position $x$,
and write $|\mathcal{T}_x|$ for the total length of the tree.
For the tree at nearby position $x + dx$ to be different,
there must have been a crossing-over between $x$ and $x + dx$
in one of the $|\mathcal{T}_x|$ meioses that gave birth to those individuals.
(Recall that these ``individuals'' are haploid.)
Since we measure distance along the genome
so that length is equal to the expected number of crossing-overs per generation,
the expected distance until the next crossing-over is $1/|\mathcal{T}_x|$.
If every crossing-over changed the marginal tree,
then this would imply, for consistency,
that the expected total number of times that the marginal tree changes along the genome
is equal to the total area of the tree sequence (as defined above).
Not every such crossing-over changes the marginal tree, but most do,
so the total area of the tree sequence,
multiplied by four,
gives an upper bound on the expected number of edges in the tree sequence
beyond those describing the leftmost tree.
Since we are considering a chromosome of length 1,
the expected total area is equal to the mean marginal tree length, as above.

If $T$ is large relative to $N$,
so that all marginal trees have a single root with high probability,
then coalescent theory tells us that
the expected total length of the branches of a marginal tree back to the most recent common ancestor
is approximately $2N\log(N)$.
Therefore, the tree sequence describing the entire population
is expected to need no more than $2N + 8N\log(N)$ edges.
Not every new edge derives from a never-before-seen node,
but the number of nodes is at most equal to the number of edges plus the sample size.
Therefore,
we would need $O(N^2)$ space to store the complete history of the simulation,
but only $O(N \log N)$ % $O(N + N \log N)$
to store the history that is relevant to the extant population.

@@

%% Above was edges; same holds for memory usage savings due to mutations
What about mutations?
Forwards-time generation of infinite-sites mutations with total mutation rate per generation $\mu$
would produce around $\mu N T$ mutations (and the same number of sites),
simply because there was a total of $N T$ meioses.
Since mutations that are retained after simplification
are precisely those that fall on the marginal trees,
the number of mutations is proportional to the total area of the tree sequence.
By the same argument as for the number of edges,
this will result in around $\mu 2 N \log(N)$ mutations.
With $N=2 \times 10^4$ and $T=10N$,
this implies that adding neutral mutations
to the simplified tree sequence reduces the number of mutations that must be generated
by a factor of 10,000.
This could result in substantial time savings,
even without considering the
computational burden of propagating mutations forwards across generations.

Since each mutation is stored only as a single row in the mutation table,
and at most one row in the site table, the space required for $M$ mutations is $O(M)$;
combined with the $O(N \log N)$ storage for edges and nodes of a simplified tree sequence,
this implies that the full results of a simulation of $N$ individuals
having $M$ mutations can be stored in $O(N \log N + M)$ space.
In simulations of whole chromosomes, there are typically a small, bounded number of mutations per chromosome per generation,
so $M$ will be $O(N \log N)$ as well.
